Data Science Applied to Ag - Final Project - ML

### 2025-04-25 code/Draft_script.qmd
d

@@

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(10:ncol(.), ~round(., 1) )) %>%
  select(-mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

```{r}
#| message: false

@@

#Export weather_monthsum.csv

@@

# Feature engineering "planting stages" and "harvest stages":

The code chunk below is to feature engineer planting and harvest stages based on the following cut-offs for planting day of the year (plant_doy) and harvest day of the year (harvest_doy):

Planting stages:
 - early: plant_doy ≤ 120
 - normal: plant_doy = 121–150
 - late: plant_doy > 150

Harvest stages:
 - early: harvest_doy ≤ 275
 - normal: harvest_doy = 276–305
 - late: harvest_doy > 305

@@

trait_avg_df <- trait_avg_df %>%
mutate(
    planting_stages = case_when(
      plant_doy <= 120                     ~ "early",
      plant_doy >= 121 & plant_doy <= 150  ~ "normal",
      plant_doy > 150                      ~ "late"
    ) %>% 
      fct_relevel("early","normal","late"),

    harvest_stages = case_when(
      harvest_doy <= 275                    ~ "early",
      harvest_doy >= 276 & harvest_doy <= 305 ~ "normal",
      harvest_doy > 305                     ~ "late"
    ) %>% 
      fct_relevel("early","normal","late")
  ) %>%
  select( -plant_doy, -harvest_doy) %>%
  ungroup()

trait_avg_df

View(trait_avg_df)

@@

>>>>

View(trait_df)

@@

View(fieldweather)

@@

fe_month_sum_wide <- fe_month_sum %>%
  pivot_longer(cols = mean_soil.ph:sum_prcp.mm) %>% 
  mutate(varname = paste0(name, "_", month_abb)) %>% 
  dplyr::select(-name, -month_abb) %>% 
  pivot_wider(names_from = varname,
              values_from = value) %>%
  # Rounding to one decimal point
  mutate(across(10:ncol(.), ~round(., 1) )) %>%
  select(-lon, -lat, -mean_soil.ph_NA, -mean_soil.om.pct_NA, -mean_soil.k.ppm_NA, -mean_soil.p.ppm_NA, -mean_dayl.s_NA, -mean_srad.wm2_NA, -mean_tmax.c_NA, -mean_tmin.c_NA, -mean_vp.pa_NA, -sum_prcp.mm_NA)

@@

@@

@@

# End of feature engineering

# Start of XGboost

# ML Workflow with XGBoost

## Step 1: Load Libraries

We begin by loading the required libraries:

```{r}
#| message: false
#| warning: false
#install.packages("xgboost") #new pacakage
library(tidymodels)   # Core framework for modeling (includes recipes, workflows, parsnip, etc.)
library(finetune)     # Additional tuning strategies (e.g., racing, ANOVA-based tuning)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(ranger)       # Fast implementation of Random Forests
library(tidyverse)    # Data wrangling and visualization
library(doParallel)   # For parallel computing (useful during resampling/tuning)
#library(caret)       # Other great library for Machine Learning 
```

The following code chunk will rename the "fe_month_sum_wide" data frame as "weather".

@@

weather <- fe_month_sum_wide
weather

@@

# ML workflow

## 1. Pre-processing

```{r weather_split}

set.seed(931735) # Setting seed to get reproducible results

weather_split <- initial_split(
  weather, 
  prop = .7, # proption of split same as previous codes
  strata = yield  # Stratify by target variable
  )

weather_split

### a. Data split

For data split, let's use **70% training / 30% testing**.

```{r weather_train}

weather_train <- training(weather_split)  # 70% of data

weather_train #This is your traing data frame

```{r weather_test}

weather_test <- testing(weather_split)    # 30% of data

weather_test

### b. Distribution of target variable

```{r distribution}
ggplot() +
  geom_density(data = weather_train, 
               aes(x = yield),
               color = "red") +
  geom_density(data = weather_test, 
               aes(x = yield),
               color = "blue")

### c. Data processing with recipe

Creating a recipe is an easy way to port your processing steps for other data sets without needing to repeat code, and also only considering the data it is being applied to.

You can find all available recipe step options here: https://tidymodels.github.io/recipes/reference/index.html

Different model types require different processing steps.\
Let's check what steps are required for an elastic net model (linear_reg). We can search for that in this link: https://www.tmwr.org/pre-proc-table #Don't need for numeric data

```{r weather_recipe}

# Create recipe for data preprocessing
weather_recipe <- recipe(yield ~ ., data = weather_train) %>% # Remove identifier columns and months not in growing season
  step_rm(
    year,       # Remove year identifier
    site,       # Remove site identifier
    matches("Jan|Feb|Mar|Dec")  # Remove non-growing season months
  ) %>%
  # Turning every factor (prev_crop, planting_stages, harvest_stages, etc.) into dummies
  step_dummy(all_nominal_predictors()) %>%
  # Imputation for any numeric NAs   
  step_impute_median(all_numeric_predictors())

weather_recipe

```{r weather_prep}
# Prep the recipe to estimate any required statistics
weather_prep <- weather_recipe %>% 
  prep()

# Examine preprocessing steps
weather_prep
```
## 2. Training

### a. Model specification

First, let's specify:\
- the **type of model** we want to train\
- which **engine** we want to use\
- which **mode** we want to use

> Elastic nets can only be run for a numerical response variable. XgBoost can be run with either numerical (regression) or categorical (classification) explanatory variable. Therefore, we have the need to specify the mode here.

XgBoost **hyperparameters**:

-   **`trees`**: The number of boosting rounds (i.e., how many trees will be added sequentially).\
-   **`tree_depth`**: Controls how deep each individual tree can grow. Deeper trees can capture more complex interactions but also increase the risk of overfitting.\
-   **`min_n`**: Minimum number of observations required in a node for it to be split. Acts as a regularization tool to prevent overly specific splits.\
-   **`learn_rate`**: Also known as `eta`, it controls how much each additional tree contributes to the overall model. Smaller values make the model more stable but require more trees.

```{r xgb_spec}
xgb_spec <- boost_tree(
  trees = tune(),
  tree_depth = tune(),  # Maximum depth of each tree
  min_n = tune(),  # Minimum samples required to split a node
  learn_rate = tune()) %>% #Specifying XgBoost as our model type, asking to tune the hyperparameters
  set_engine("xgboost") %>% #specify engine 
  set_mode("regression")  # Set to mode
   # Total number of boosting iterations

        # Step size shrinkage for each boosting step

xgb_spec

### b. Cross-validation setup

We use 5-fold cross-validation to evaluate model performance during tuning:

set.seed(235) #34549

resampling_foldcv <- vfold_cv(weather_train, 
                              v = 5)

resampling_foldcv

resampling_foldcv$splits[[1]]

```
### c. Hyperparameter grid with Latin Hypercube Sampling

We use Latin hypercube sampling to generate a diverse grid of hyperparameter combinations:

```{r }

xgb_grid <- grid_latin_hypercube(
  tree_depth(),
  min_n(),
  learn_rate(),
  trees(),
  size = 10
)

xgb_grid
```

```{r}
ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate), #coloring the bubbles based on learn_rate
                 size = trees), #size of the bubbles are based on the tress
             alpha = .5,
             show.legend = FALSE)
```

## 3. Model Tuning

```{r xgb_grid_result}

#install.packages("doParallel")
#install.packages("parallel")

library(doParallel)
library(parallel)

set.seed(76544)

#parallel processing
registerDoParallel(cores = parallel::detectCores()-1)

xgb_res <- tune_race_anova(object = xgb_spec,
                      preprocessor = weather_recipe,
                      resamples = resampling_foldcv,
                      grid = xgb_grid,
                      control = control_race(save_pred = TRUE))

stopImplicitCluster()

beepr::beep()

xgb_res$.metrics[[2]]
